%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Contents: The Bibliography
% File: biblio.tex (lshort2e.tex)
% $Id$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{99}
\addcontentsline{toc}{chapter}{\bibname}
\bibitem{manual} Leslie Lamport.  \newblock \emph{{\LaTeX:} A Document
    Preparation System}.  \newblock Addison-Wesley, Reading,
  Massachusetts, second edition, 1994, ISBN~0-201-52983-1.

\bibitem{texbook} Donald~E. Knuth.  \newblock \textit{The \TeX{}book,}
  Volume~A of \textit{Computers and Typesetting}, Addison-Wesley,
  Reading, Massachusetts, second edition, 1984, ISBN~0-201-13448-9.

\bibitem{companion} Frank Mittelbach, Michel Goossens, Johannes Braams,
  David Carlisle, Chris Rowley.  \newblock \emph{The {\LaTeX} Companion, (2nd Edition)}.  \newblock
  Addison-Wesley, Reading, Massachusetts, 2004, ISBN~0-201-36299-6.

\bibitem{graphicscompanion} Michel Goossens, Sebastian Rahtz and Frank
  Mittelbach.  \newblock \emph{The {\LaTeX} Graphics Companion}.  \newblock
  Addison-Wesley, Reading, Massachusetts, 1997, ISBN~0-201-85469-4.

\bibitem{local} Each \LaTeX{} installation should provide a so-called
  \emph{\LaTeX{} Local Guide}, which explains the things that are
  special to the local system.  It should be contained in a file called
  \texttt{local.tex}. Unfortunately, some lazy sysops do not provide such a
  document. In this case, go and ask your local \LaTeX{} guru for help.

\bibitem{usrguide} \LaTeX3 Project Team.  \newblock \emph{\LaTeXe~for
    authors}.  \newblock Comes with the \LaTeXe{} distribution as
  \texttt{usrguide.tex}.

\bibitem{clsguide} \LaTeX3 Project Team.  \newblock \emph{\LaTeXe~for
    Class and Package writers}.  \newblock Comes with the \LaTeXe{}
  distribution as \texttt{clsguide.tex}.

\bibitem{fntguide} \LaTeX3 Project Team.  \newblock \emph{\LaTeXe~Font
    selection}.  \newblock Comes with the \LaTeXe{} distribution as
  \texttt{fntguide.tex}.

\bibitem{graphics} D.~P.~Carlisle.  \newblock \emph{Packages in the
    `graphics' bundle}.  \newblock Comes with the `graphics' bundle as
  \texttt{grfguide.tex}, available from the same source your \LaTeX{}
  distribution came from.

\bibitem{verbatim} Rainer~Sch\"opf, Bernd~Raichle, Chris~Rowley.
\newblock \emph{A New Implementation of \LaTeX's verbatim
  Environments}.
 \newblock Comes with the `tools' bundle as
  \texttt{verbatim.dtx}, available from the same source your \LaTeX{}
  distribution came from.

\bibitem{cyrguide} Vladimir Volovich, Werner Lemberg and \LaTeX3 Project Team.
    \newblock \emph{Cyrillic languages support in \LaTeX}.
    \newblock Comes with the \LaTeXe{} distribution as
  \texttt{cyrguide.tex}.

\bibitem{catalogue} Graham~Williams.  \newblock \emph{The TeX
    Catalogue} is a very complete listing of many \TeX{} and \LaTeX{}
    related packages.
  \newblock Available online from \CTAN|help/Catalogue/catalogue.html|

\bibitem{xy-pic} Kristoffer H. Rose.
  \newblock \emph{\Xy-pic User's Guide}.  \newblock
  Downloadable from CTAN with \Xy-pic distribution

\bibitem{metapost} John D. Hobby.
  \newblock \emph{A User's Manual for \MP}. \newblock
  Downloadable from \url{http://cm.bell-labs.com/who/hobby/}

\bibitem{unbound} Alan Hoenig.
  \newblock \emph{\TeX{} Unbound}. \newblock Oxford University Press, 1998,
    ISBN 0-19-509685-1; 0-19-509686-X (pbk.)

\bibitem{ursoswald} Urs Oswald.
    \newblock \emph{Graphics in \LaTeXe{}}, containing some Java source files for
    generating arbitrary circles and ellipses within the \texttt{picture} environment,
    and \emph{\MP{} - A Tutorial}.
  \newblock Both downloadable from \url{http://www.ursoswald.ch}

\bibitem{pgfplot} Till Tantau.
  \newblock \emph{TikZ\&PGF Manual}.\newblock
  Download from \CTAN|graphics/pgf/base/doc/generic/pgf/pgfmanual.pdf|

% new items for XeLaTeX
\bibitem{polyglossia} Fran\c{c}ois Charette.
    \newblock \emph{Polyglossia: A Babel Replacement for \hologo{XeLaTeX}}.
    \newblock Comes with the \TeX Live distribution as
  \texttt{polyglossia.pdf}. (Type \texttt{texdoc polyglossia} on the command line.)

\bibitem{arabxetex} Fran\c{c}ois Charette.
    \newblock \emph{An Arab\TeX-like interface for typesetting languages
     in Arabic script with \hologo{XeLaTeX}}.
    \newblock Comes with the \TeX Live distribution as
  \texttt{arabxetex.pdf}. (Type \texttt{texdoc arabxetex} on the command line.)

\bibitem{fontspec} Will Robertson and Khaled Hosny.
    \newblock \emph{The \texttt{fontspec} package}.
    \newblock Comes with the \TeX Live distribution as
  \texttt{fontspec.pdf}. (Type \texttt{texdoc fontspec} on the command line.)

\bibitem{xgreek} Apostolos Syropoulos.
    \newblock \emph{The \texttt{xgreek} package}.
    \newblock Comes with the \TeX Live distribution as
  \texttt{xgreek.pdf}. (Type \texttt{texdoc xgreek} on the command line.)

\bibitem{bidi} Vafa Khalighi.
    \newblock \emph{The \texttt{bidi} package}.
    \newblock Comes with the \TeX Live distribution as
  \texttt{bidi.pdf}. (Type \texttt{texdoc bidi} on the command line.

\bibitem{xepersian} Vafa Khalighi.
    \newblock \emph{The \texttt{XePersian} package}.
    \newblock Comes with the \TeX Live distribution as
  \texttt{xepersian-doc.pdf}. (Type \texttt{texdoc xepersian} on the command line.

\bibitem{xecjk} Wenchang Sun.
    \newblock \emph{The \texttt{xeCJK} package}.
    \newblock Comes with the \TeX Live distribution as
  \texttt{xeCJK.pdf}. (Type \texttt{texdoc xecjk} on the command line.

\end{thebibliography}


%

% Local Variables:
% TeX-master: "lshort2e"
% mode: latex
% mode: flyspell
% End:
