%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Contents: Who contributed to this Document
% $Id$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{small} 
  \noindent Copyright \copyright 1995-2016 Tobias Oetiker and Contributors.  All rights reserved.
 
  This document is free; you can redistribute it and/or modify it
  under the terms of the GNU General Public License as published by
  the Free Software Foundation; either version 2 of the License, or
  (at your option) any later version.
  
  This document is distributed in the hope that it will be useful, but
  \emph{without any warranty}; without even the implied warranty of
  \emph{merchantability} or \emph{fitness for a particular purpose}\@.  See the GNU
  General Public License for more details.
  
  You should have received a copy of the GNU General Public License
  along with this document; if not, write to the Free Software
  Foundation, Inc., 675 Mass Ave, Cambridge, MA 02139, USA.

\end{small}

\chapter{Thank you!}
\noindent Much of the material used in this introduction comes from an
Austrian introduction to \LaTeX\ 2.09 written in German by:
\begin{verse}
\contrib{Hubert Partl}{partl@mail.boku.ac.at}%
{Zentraler Informatikdienst der Universit\"at f\"ur Bodenkultur Wien}
\contrib{Irene Hyna}{Irene.Hyna@bmwf.ac.at}%
   {Bundesministerium f\"ur Wissenschaft und Forschung Wien}
\contrib{Elisabeth Schlegl}{no email}%
   {in Graz}
\end{verse}

If you are interested in the German document, you can find a version
updated for \LaTeXe{} by J\"org Knappen at
\CTAN|info/lshort/german|

\newpage \noindent The
following individuals helped with corrections, suggestions and
material to improve this paper. They put in a big effort to help me
get this document into its present shape. I would like to
sincerely thank all of them. Naturally, all the mistakes you'll find
in this book are mine. If you ever find a word that is spelled
correctly, it must have been one of the people below dropping me a
line.

If you want to contribute to this booklet, you can find all the source code
on \url{https://github.com/oetiker/lshort}. Your pull requests will be
appreciated.

{\flushleft\small
Eric~Abrahamsen,        % <eric@ericabrahamsen.net>
Lenimar~Nunes~de~Andrade, % <lenimar@mat.ufpb.br> 12 Nov 1999
Eilinger~August,        % <eaugust@student.ethz.ch>
Rosemary~Bailey,        % <r.a.bailey@qmw.ac.uk> 0.2
Barbara~Beeton,         % <bnb@ams.org>
Marc~Bevand,            % <bevand_m@epita.fr>
Connor~Blakey,          % it's Ligatures!
Salvatore~Bonaccorso,   % <bonaccos@ee.ethz.ch>
Pietro~Braione,         % <braione@elet.polimi.it>
Friedemann~Brauer,      % <fbrauer@is.dal.ca> 3.4
Markus~Br\"uhwiler,     % <m.br@switzerland.org>
Jan~Busa,               % <busaj@ccsun.tuke.sk>
David~Carlisle,         % GONE <carlisle@cs.man.ac.uk> 1.0
Neil~Carter,            % <n.carter@Swansea.ac.uk>
Carl~Cerecke,           % <cdc@cosc.canterbury.ac.nz>
Mike~Chapman,           % <chapman@eeh.ee.ethz.ch> 3.16
Pierre~Chardaire,       % <pc@sys.uea.ac.uk>
Xingyou~Chen,           % <niatlantice@gmail.com> 5.04
Christopher~Chin,       % <chris.chin@rmit.edu.au> 3.1
Diego~Clavadetscher,    % <dc@clavatax.ch>
Wim~van~Dam,            % GONE <wimvdam@cs.kun.nl> 2.2
Benjamin~Deschwanden    % <vdeschwb@student.ethz.ch>
Jan~Dittberner,         % <jan@jan-dittberner.de> 3.15
Michael~John~Downes,    % <mjd@ams.org> 14 Oct 1999
Matthias~Dreier,        % <dreier@ostium.ch>
David~Dureisseix,       % <dureisse@lmt.ens-cachan.fr> 1.1
Hans~Ehrbar,            % <ehrbar@econ.utah.edu>
Elliot,                 % GONE <enh-a@minster.york.ac.uk> 1.1
Rockrush~Engch,         % <niatlantice@gmail.com>
William~Faulk,          % <wfaulk@webassign.net>
Robin~Fairbairns,       % <robin.fairbairns@cl.cam.ac.uk> 0.2 1.0
Johan~Falk,             % <johan@vaxjonexus.com> 5.0.1
J\"org~Fischer,         % <j.fischer@xpoint.at> 3.16
Frank~Fischli,          % <fischlifaenger@gmx.ch>
Daniel~Flipo,           % <daniel.flipo@univ-lille1.fr>
Frank,                  % <frank@freezone.co.uk> 11 Feb 2000
Mic~Milic~Frederickx,   % <mic.milic@web.de>
David~Frey,             % <david@eos.lugs.ch> 2.2
Erik~Frisk,             % <frisk@isy.liu.se> 3.4
Hans~Fugal,             % <hans@fugal.net>
Robert~Funnell,         % <robert.funnell@mcgill.ca> 5.1
Greg~Gamble,            % <gregg@maths.uwa.edu.au> 2.2
Andy~Goth,              % <unununium@openverse.com>
Cyril~Goutte,           % <goutte@ei.dtu.dk> 2.1 2.2
Kasper~B.~Graversen,    % <kbg@dkik.dk>
Arlo~Griffiths,         % <a.griffiths@let.leidenuniv.nl>
Alexandre~Guimond,      % <guimond@iro.umontreal.ca> 0.9
Neil~Hammond,           % <nfh@dmu.ac.uk> 0.3
Christoph~Hamburger,    % <ch.hamburger@gmail.com>
Rasmus~Borup~Hansen,    % GONE <rbhfamos@math.ku.dk> 0.2 0.9 0.91 0.92 1.9.9
Joseph~Hilferty,        % <hilferty@fil.ub.es>
Daniel~Hirsbrunner,     % <dhirsbrunner1@gmail.com>
Martien~Hulsen,         % <m.a.hulsen@Wbmt.tudelft.nl> 1.0 1.1
Bj\"orn Hvittfeldt,     % <bjorn@hvittfeldt.com> 3.13
Morten~H\o gholm,       % <morten.hoegholm@latex-project.org>
Werner~Icking,          % <werner.icking@gmd.de> 3.1
Eric~Jacoboni,          % GONE <jacoboni@enseeiht.fr> 0.1 0.9
Jakob,                  % <diness@get2net.dk>
Alan~Jeffrey,           % <alanje@cogs.sussex.ac.uk> 0.2
Martin~Jenkins,         % xqp.ltd@gmail.com 5.04
Byron~Jones,            % <bj@dmu.ac.uk> 1.1
David~Jones,            % GONE <djones@ca.mcmaster.dcss.insight> 1.1
Johannes-Maria~Kaltenbach, % <kaltenbach@zeiss.de> 3.01
Nils~Kanning,           % <nils@kanning.de>
Andrzej~Kawalec,        % GONE <akawalec@prz.rzeszow.pl> 1.9.9
Christian~Kern,         % <ck@unixen.hrz.uni-oldenburg.de> 2.1
Alain~Kessi,            % <alain_kessi@hotmail.com> 2.2
Axel~Kielhorn,          % <a.kielhorn@web.de>
Sander~de~Kievit,       % <Skievit@ucu.uu.nl>
Kjetil~Kjernsmo,        % <kjetil.kjernsmo@astro.uio.no> 3.2
Tobias~Klauser,		% <tklauser@access.unizh.ch> 4.17
J\"org~Knappen,         % <knappen@vkpmzd.kph.uni-mainz.de> 0.1
Michael~Koundouros,     % <mkoundouros@hotmail.com>
Matt~Kraai,             % <matt.kraai@amo.abbott.com>
Tobias~Krewer,          % <tobias.krewer@googlemail.com>
Flori~Lambrechts,       % <f.lambrechts@softhome.net>
Mike~Lee,               % <rmrstar@gmail.com>
Maik~Lehradt,           % <greek@uni-paderborn.de> 0.1
R\'emi~Letot,           % <r_letot@yahoo.com>
Axel~Liljencrantz,	% <axel.liljencrantz@byv.kth.se>
Jasper~Loy,             % <jasper.loy@gmail.com>
Johan~Lundberg,         % <p99jlu@physto.se>
Martin~Maechler,        % <maechler@stat.math.ethz.ch> 2.2
Alexander~Mai,          % <alexander.mai@physik.tu-darmstadt.de> 3.8
Claus~Malten,           % GONE <asi138%bitnet.djukfa11@bitnet.cearn> 1.1
Kevin~Van~Maren,        % <vanmaren@fast.cs.utah.edu> 24 Nov 1999
Pablo~Markin,
I.~J.~Vera~Mar\'un,     % <i.j.veramarun@ewi.utwente.nl>
Hendrik~Maryns,         % <hendrik.maryns@ugent.be>
Chris~McCormack,        % GONE <chrismc@eecs.umich.edu> 0.1
Aleksandar~S.~Milosevic, % <aleksandar.milosevic@yale.edu>
Henrik~Mitsch,          % <henrik.mitsch@gmx.at>
Stefan~M.~Moser,        % <stefan.moser@ieee.org>
Armin~M\"uller,		% arm.in@web.de
Philipp~Nagele,         % <philipp.nagele@t-systems.com>
Richard~Nagy,           % <r.nagy@nameshield.net>
Manuel~Oetiker,         % <manuel@oetiker.ch>
Urs~Oswald,             % <osurs@bluewin.ch>
Hubert~Partl,           % <partl@mail.boku.ac.at> 0.2 1.1
Marcelo~Pasin,          % <pasin@di.fc.ul.pt>
Martin~Pfister,		% <m@rtinpfister.ch>
Lan~Thuy~Pham,          % <lan.thuy.pham@gmail.com>
Breno~Pietracci,        % <bpietracci@gmail.com>
Demerson~Andre~Polli,   % <polli@linux.ime.usp.br>
Maksym~Polyakov,        % <polyama@myrealbox.com>
Nikos~Pothitos,		% <n.pothitos@di.uoa.gr>
John~Refling,           % <refling@sierra.lbl.gov> 0.1 0.9
Mike~Ressler,           % <ressler@cougar.jpl.nasa.gov> 0.1 0.2 0.9 1.0 1.9.9
Brian~Ripley,           % <ripley@stats.ox.ac.uk> 2.1
Kurt~Rosenfeld,		% <kurt@isis.poly.edu>
Bernd~Rosenlecher,      % <9rosenle@informatik.uni-hamburg.de> 10 Feb 2000
Chris~Rowley,           % <c.a.rowley@open.ac.uk> 0.91
Young~U.~Ryu,           % <ryoung@utdallas.edu> 2.1
Risto~Saarelma,         % <risto.saarelma@cs.helsinki.fi>
Andr{\'a}s~Salamon,     % <andras.salamon@comlab.ox.ac.uk>
Jos\'e~Carlos~Santos,   % <jcsantos@fc.up.pt>
Christopher~Sawtell,    % <csawtell@xtra.co.nz> 1 Sep 1999
Gilles~Schintgen,       % <gschintgen@internet.lu>
Craig~Schlenter,        % <cschle@lucy.ee.und.ac.za> 0.1 0.2 0.9
Hanspeter~Schmid,       % <schmid@isi.ee.ethz.ch>
Baron~Schwartz,         % <bps7j@cs.virginia.edu>
Jordi~Serra~i~Solanich, % <solanich@gmail.com>
Miles~Spielberg,        % <zeibach@hotmail.com>
Susan~Stewart,
Matthieu~Stigler,
Geoffrey~Swindale,      % <geofftswin@ntlworld.com>
Laszlo~Szathmary,       % <szathml@delfin.klte.hu>
Boris~Tobotras,         % <tobotras@jet.msk.su>
Josef~Tkadlec,          % <tkadlec@math.feld.cvut.cz> 2.0 2.2
Scott~Veirs,            % <scottv@ocean.washington.edu>
Didier~Verna,           % <verna@inf.enst.fr> 2.2
Carl-Gustav~Werner,     % <carl-gustav.werner@math.lu.se> 11 Oct 1999, 3.16
Fabian~Wernli,          % <wernli@iap.fr> 3.2
Matthew~Widmann,        % <mtwdmn@gmail.com>
David~Woodhouse,        % <dwmw2@infradead.org> 3.16
Chris~York,             % <c.s.york@Cummins.com> 21 Nov 1999
Rick~Zaccone,           % <zaccone@bucknell.edu> 2.2
Fritz~Zaucker,          % <zaucker@ee.ethz.ch> 3.0
and Mikhail~Zotov.      % <zotov@eas.npi.msu.su> 3.1
}

\vspace*{\stretch{1}}

\pagebreak
\endinput
%

% Local Variables:
% TeX-master: "lshort2e"
% mode: latex
% mode: flyspell
% End:
